---
title: "Primary Healthcare Andalucia"
output: html_notebook
---
The purpose of this project is comparing the primary healthcare among the eight provinces in Andalucía, so as to verify if there are differences within the same region (Comunidad Autónoma - CA)

According to a survey carried out by Deloitte, 85% of the health professional questioned considered that there are differences among centers within the same CA. 

It can be only challenge, however, the primary healthcare sector, where the information is broken down by small areas. The data related to hospital is only collected by CA. 

There are, then, three parameters to meassure the resources given to each area: The amount of healthcare centers, the quantity of sanitary professionals and the frequency. 

# Healthcare centers

## Centers

First, I upload the file that contains all the healthcare centers in Spain and their basic areas. 

```{r}
centersAT <- read.csv("20171208_Centros_de_Atencion_Primaria.csv", sep = ",", stringsAsFactors = FALSE)
View(centersAT)
```

I change the columns title I am interested in.

```{r}
names(centersAT) <- gsub("Zona.B..sica", "zona", names(centersAT))
names(centersAT) <- gsub("Comunidad.Aut..noma", "CA", names(centersAT))
names(centersAT) <- gsub("C..digo.Postal", "CP", names(centersAT))
names(centersAT) <- gsub("X..rea.de.Salud", "AreaSalud", names(centersAT))
names(centersAT) <- gsub("Modalidad.Gesti..n", "Modalidad Gestion", names(centersAT))
names(centersAT) <- gsub("Dependencia.de.Gesti..n", "Dependencia", names(centersAT))
View(centersAT)
```

Next I check if there is some rows wrong classed. For example, if I have other names than the 17 (plus Ceuta and Melilla) Spanish regions (Comunidades Autónomas) in that column. With table() I can see which names do not belong to that column type.

```{r}
table(centersAT$CA)
```
The names may cause problems in the future. I fix those that look weird, avoiding non common characters in English. 

I could do it only for Andalucía, but, if I (or anyone) will need to use it again, we'll already have work done. 

```{r}
centersAT$CA[centersAT$CA == "ANDALUC\303\215A"] <- "ANDALUCIA"
centersAT$CA[centersAT$CA == "CASTILLA Y LE\303\223N"] <- "CASTILLA Y LEON"
centersAT$CA[centersAT$CA == "MURCIA (REGI\303\223N DE)"] <- "MURCIA"
centersAT$CA[centersAT$CA == "ARAG\303\223N"] <- "ARAGON"
centersAT$CA[centersAT$CA == "CATALU\303\221A"] <- "CATALUNYA"
centersAT$CA[centersAT$CA == "PA\303\215S VASCO"] <- "PAIS VASCO"
View(centersAT)
```

I am going now to create a subset only with Andalucía to clean the data that I am going to use. 

```{r}
atAND <- subset(centersAT, centersAT$CA == "ANDALUCIA")
View(atAND)
```

I need now to repeat the same process with "provincias" (smaller division), "Area Salud" and "zona" (smaller subdivisions) and "Tipo de Centro", as these are the parameter I am going to use. 

```{r}
table(atAND$Provincia)
```
```{r}
atAND$Provincia[atAND$Provincia == "ALMER\303\215A"] <- "ALMERIA"
atAND$Provincia[atAND$Provincia == "M\303\201LAGA"] <- "MALAGA"
atAND$Provincia[atAND$Provincia == "C\303\201DIZ"] <- "CADIZ"
atAND$Provincia[atAND$Provincia == "C\303\223RDOBA"] <- "CORDOBA"
atAND$Provincia[atAND$Provincia == "JA\303\211N"] <- "JAEN"
View(atAND)
```

By using table(), it also counts the number of times that a value appears, which means how many centers there are in that area.

However, there are two types of centers.
```{r}
table(atAND$Tipo.de.Centro)
```
With table(atAND$Area Salud), it prints how many centers by small territorial division. 
```{r}
table(atAND$`Area Salud`)
```
I identify that \303\ means accents, 201 = A, 211 = E, 232 = U, 223 = O, 215 = I, 221 = Ñ, 
```{r}
newlist <- atAND$`Area Salud`
newlist <- gsub("\303\211","E",gsub("\303\201","A",gsub("\303\215","I", gsub("\303\223", "O", gsub("\303\232","U", gsub("\303\221","N", newlist))))))
table(newlist)
#Store the function to reuse later
formulnames <- function(newlist) { gsub("\303\211","E",gsub("\303\201","A",gsub("\303\215","I", gsub("\303\223", "O", gsub("\303\232","U", gsub("\303\221","N", newlist))))))}
atAND$areasaludlimpio <- newlist
```

By using table(atAND$Area Salud), it prints the numbers of centers per small area. I could keep this value to know which one has the largest number. However, I would prefer comparing ratio. For that purpose, I'm going to use another file with the population in each small area.

But first, I make another basic table with only the information I am interested in.

```{r}
atBasic <- data.frame(atAND$Provincia, atAND$areasaludlimpio, atAND$Tipo.de.Centro)
colnames(atBasic) <- c("Provincia", "AreaSalud", "Tipo")
#Keeping only the names in the "Área Salud" column
atBasic$AreaSalud <- gsub("^.*: ", "", atBasic$AreaSalud)
atBasic
```


## Population

I upload the file and make a subset only with the data from 2016 to compare with the information above. 

```{r}
population <- read.csv2("poblacionasignada.csv", stringsAsFactors = FALSE)
pop2016 <- subset(population, population$A.o == "2016")
colnames(pop2016) <- c("Ano", "Provincia", "AreaSalud", "Grupos", "Numero")
```

I clean again the names

```{r}
table(pop2016$Provincia)
```
```{r}
pop2016$Provincia[pop2016$Provincia == "ALMER\315A"] <- "ALMERIA"
pop2016$Provincia[pop2016$Provincia == "C\301DIZ"] <- "CADIZ"
pop2016$Provincia[pop2016$Provincia == "C\323RDOBA"] <- "CORDOBA"
pop2016$Provincia[pop2016$Provincia == "JA\311N"] <- "JAEN"
pop2016$Provincia[pop2016$Provincia == "M\301LAGA"] <- "MALAGA"
View(pop2016)
```

And I install the package "dplyr"

```{r}
install.packages("dplyr")
library(dplyr)
```

I have too much detail. I am interested in the total population allocated in each small area. 

```{r}
pop2016 %>% group_by(AreaSalud) %>% summarise(Numero=sum(Numero))
#I store it because I will need it later
pop2016Total <- pop2016 %>% select(Provincia, AreaSalud, Numero) %>% group_by(AreaSalud) %>% summarise(Numero=sum(Numero))
#Cleaning names
formulnames(pop2016Total$AreaSalud)
pop2016Total$AreaSalud
```
Some names need more cleaning
```{r}
pop2016Total$AreaSalud <- gsub("\341", "a", gsub("\363", "o", gsub("\355", "i", gsub("\351", "e", gsub("\361", "n", pop2016Total$AreaSalud)))))
pop2016Total$AreaSalud <- toupper(pop2016Total$AreaSalud)
pop2016Total$AreaSalud
```

## Counting

<strong>The most and the least</strong>

I group by Área Salud to know the quantity of centers (no matter the type) in each Área Salud. 

```{r}
atBasic %>% select(Provincia, AreaSalud, Tipo) %>% group_by(Provincia, AreaSalud) %>% summarise(Tipo = n()) %>% arrange(Tipo)
#I will use it later to calculate the ratio
centersTotal <- atBasic %>% select(Provincia, AreaSalud, Tipo) %>% group_by(Provincia, AreaSalud) %>% summarise(Tipo = n()) %>% arrange(Tipo)
```
```{r}
centersTotal %>% select(AreaSalud, Tipo) %>% group_by(AreaSalud) %>% summarise(Total = sum(Tipo)) %>%  arrange(Total)
```

The top three is:
- Metropolitano de Granada
- A.G.S. Norte de Jaén
- Almería

And at the bottom:
- Costa del sol
- Córdoba and Condado - Campina
- Valle del guadalhorce

```{r}
View(centersTotal)
```


<strong>By type</strong>
I would like to know how many centers of each type are by area, because a "centro de salud" is bigger and with more resources than "consultorio local".

```{r}
atBasic %>% select(Provincia, AreaSalud, Tipo) %>% filter(Tipo == "CENTRO SALUD")  %>% group_by(AreaSalud) %>% summarise(Tipo = n()) %>% arrange(Tipo)

```
By "centro de salud", Sevilla and Málaga are at the top and A.G.S Norte de Málaga and A.G.S. Serranía de Málaga and A.G.S. Norte de Córdoba at the bottom. 

Regarding "consultorio local", Metropolitano de Granada and A.G.S. Norte de Jaén are the top two and Sevilla and Granada at the bottom. 

```{r}
atBasic %>% select(Provincia, AreaSalud, Tipo) %>% filter(Tipo == "CONSULTORIO LOCAL")  %>% group_by(AreaSalud) %>% summarise(Tipo = n()) %>% arrange(Tipo)
```
```{r}
#Filtering by province instead of by health area
atBasic %>% select(Provincia, Tipo) %>% filter(Tipo == "CONSULTORIO LOCAL")  %>% group_by(Provincia) %>% summarise(Tipo = n()) %>% arrange(Tipo)
#downloading to visualize
write.csv(atBasic, file = "atBasic.csv")
```
```{r}
atBasic %>% select(Provincia, Tipo) %>% filter(Tipo == "CENTRO SALUD")  %>% group_by(Provincia) %>% summarise(Tipo = n()) %>% arrange(Tipo)
```


## Comparing

<strong>People by healthcare center</strong>

To have the ratio, I would need to combine both table: centersTotal and pop2016Total. 

I will first merge both by "AreaSalud", so I need to clean both table first, so as the names match. But, the first time I did it, I got a table with less rows than the original one. I checked again the spelling and did more cleaning. 

```{r}
centersTotal[order(centersTotal$AreaSalud),]
```


```{r}
centersTotal[28,] 
centersTotal[28,1] <- "HUELVA"
centersTotal[28,2] <- "A.G.S. NORTE DE HUELVA"
centersTotal[order(centersTotal$AreaSalud),]
```

```{r}
centersTotal$AreaSalud <- gsub("-", " ", centersTotal$AreaSalud)
centersTotal
```
```{r}
centersTotal[25,1] <- "MALAGA"
centersTotal[25,2] <- "A.G.S. ESTE DE MALAGA AXARQUIA"
centersTotal[25,]
```
```{r}
centersTotal
```

```{r}
centersTotal[14,1] <- "CADIZ"
centersTotal[14,2] <- "BAHIA DE CADIZ LA JANDA"
centersTotal[14,]
```

```{r}
centersTotal[11,1]<- "MALAGA"
centersTotal[11,]
```

```{r}
centersTotal[6,2] <- "CONDADO CAMPINA"
centersTotal[6,]
```


```{r}
pop2016Total[order(pop2016Total$AreaSalud),]
```

```{r}
pop2016Total$AreaSalud <- gsub("-", " ", pop2016Total$AreaSalud)
pop2016Total
```

```{r}
pop2016Total[2,1] <- "A.G.S. ESTE DE MALAGA AXARQUIA"
pop2016Total[2,]
```

```{r}
pop2016Total[7,]
```


```{r}
ratioAT <- merge(pop2016Total, centersTotal, "AreaSalud")
ratioAT
```


```{r}
centersTotal[order(centersTotal$AreaSalud),]
```

<strong>Calculating ratio</strong>
I calculate ratio per "provincia"

```{r}
ratioATProv <- select(ratioAT, Numero, Provincia, Tipo) %>% group_by(Provincia) %>% summarise(Numero = sum(Numero), Tipo = sum(Tipo))
ratioATProv
```
```{r}
ratioATProv <- ratioATProv %>% mutate(ratio_centersProv = Numero / Tipo) %>% arrange(ratio_centersProv)
ratioATProv
```
```{r}
ratioATProv %>% summarise(mean(ratio_centersProv))
```

```{r}
ratioATProv <- ratioATProv %>% mutate(differenceProv = (ratio_centersProv - mean(ratio_centersProv)) / mean(ratio_centersProv)) 
ratioATProv
#download to visualize later
write.csv(ratioATProv, "ratioATProv.csv")
```
How big is the difference between the first and the last?

```{r}
(21402.116 - 2742.694)/2742.694
```


I want to calculate how many people by center in each area.
```{r}
ratioATArea <- select(ratioAT, AreaSalud, Numero, Tipo) %>% group_by(AreaSalud) %>% summarise(Numero = sum(Numero), Tipo = sum(Tipo))
ratioATArea
```

```{r}
ratioATArea <- ratioATArea %>% mutate(ratio_centers = Numero / Tipo) %>% arrange(ratio_centers)
ratioATArea
write.csv(ratioATArea, "ratioATArea.csv")
```

```{r}
ratioATArea %>% summarise(mean(ratio_centers), median(ratio_centers), sd(ratio_centers))
```

The differences are quite prominent. In A.G.S. Norte de Huelva there are 1,132.7 people by center, while in Costa del Sol there are 19,999.4 people by center. 

How much population are there in each area in relation to the first one?
```{r}
ratioATArea %>% mutate(difference = (ratio_centers - 1132.750)/1132.750) %>% arrange(ratio_centers)
```
The difference between the first and the last is 1665%, what means that each Healthcare center in Costa del Sol area has 16.6 more times people allocated than each center in the area A.G.S. Norte de Huelva. 

And, how much population in relation to the mean?
```{r}
ratioATArea %>% mutate(difference = (ratio_centers - mean(ratio_centers))/mean(ratio_centers)) %>% arrange(ratio_centers)
```

## Emergency non in hospital

There is a part of the emergency attendance which belongs to the Primary Healthcare. This activity is carried out in centers different from the hospitals. 

This information is not separated by "Area de Salud", but "Provincias".

```{r}
emergency <- read.csv("dispositivosurgextrahospital.csv")
emergency$Provincia <- formulnames(emergency$Provincia)
```

```{r}
emergencyAND <- subset(emergency, emergency$Provincia == "GRANADA" | emergency$Provincia == "JAEN" | emergency$Provincia == "CORDOBA" | emergency$Provincia == "MALAGA" | emergency$Provincia == "ALMERIA" | emergency$Provincia == "CADIZ" | emergency$Provincia == "HUELVA" | emergency$Provincia == "SEVILLA")
#This is the table I am going to use
emergencyANDTotal <- emergencyAND %>% select(Provincia, Nombre) %>% group_by(Provincia) %>% summarise(Nombre = n()) %>% arrange(Nombre)
emergencyANDTotal
```
Ratio usign population allocated in this "provincia"
```{r}
pop2016$Provincia <- formulnames(pop2016$Provincia)
pop2016Provincia
```
```{r}
pop2016Provincia <- pop2016 %>% select(Provincia, Numero) %>% group_by(Provincia) %>% summarise(Numero=sum(Numero, na.rm = TRUE))
pop2016Provincia
```

```{r}
ratioEmergency <- merge(pop2016Provincia, emergencyANDTotal, "Provincia")
ratioEmergency <- ratioEmergency %>% mutate(ratio_emergency = Numero / Nombre) %>% arrange(ratio_emergency)
ratioEmergency
write.csv(ratioEmergency, "ratioER.csv")
```

```{r} 
ratioEmergency %>% mutate(ratio_emergency = Numero / Nombre) %>% summarise(mean = mean(ratio_emergency), median = median(ratio_emergency), sd = sd(ratio_emergency))
``` 
```{r}
ratioEmergency %>% mutate(difference = (ratio_emergency - 12525.61) / 12525.61) %>% arrange(difference)
```
How big the last related to the first?
```{r}
(33640.76 - 12525.61)/12525.61
```


